Revisiting Short-Chain Fatty Acids and the Microbiota in Colorectal Cancer

### 2018-05-22 submission/manuscript.Rmd
add complete result section template

**Decreased SCFA concentrations are not associated with tumors.** We used frozen fecal samples from `r length(metai$sample)` individuals and HPLC to measure acetate, butyrate, and propionate concentrations at a cross-sectional point in time. There was no difference between individuals with normal colons and patients with either an adenoma or carcinoma for any of the SCFAs measured after multiple comparison correction (P-value `r ifelse(min(pull(hplc_groups_scfa_kruskal, bh)) > 0.15, invisible(">"), invisible("<"))` 0.15) [Figure 1A - 1C]. We next measured the concentration of SCFAs in `r length(metaf$initial)` patients with an adenoma or carcinoma in which we had pre- and post-treatment fecal samples. Although there was a general trend for increasing acetate, butyrate, and propionate concentrations after treatment for tumors, there was no significant difference pre- and post-treatment for patients with adenoma (P-value `r ifelse(min(filter(hplc_treat_scfa_kruskal, disease == "adenoma") %>% pull(bh)) > 0.20, invisible(">"), invisible("<"))` 0.20) or carcinoma (P-value `r ifelse(min(filter(hplc_treat_scfa_kruskal, disease == "carcinoma") %>% pull(bh)) > 0.80, invisible(">"), invisible("<"))` 0.80) [Figure 1D - 1F]. Even though there was no change in SCFA concentrations between individuals with normal colons and those with tumors, this information could still be important to help classify disease.

**Random Forest models with SCFA concentrations do not classify tumors better.** Using the Random Forest algorithm we built models to classify normal versus adenoma and normal versus carcinoma with OTUs or OTUs and SCFA concentrations. For both adenoma and carcinoma models, there was no difference between the median AUC of models with or without SCFA concentrations (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05) [Figure 2]. Although there was no added information gained for classification of disease by including SCFAs to Random Forest models, it is possible that the genes for enzymes involved in SCFA pathways may change due to disease status; where a smaller number of microbes are responsible for the observed SCFA concentrations.

**Changes in genes for enzymes involved in SCFA synthesis are not associated with tumors.** Using a list of specific genes that are important for the synthesis of SCFAs [Table S1], we looked for differences in gene abundance based on individuals having normal colons, adenomas, or carcinomas. First, using imputed gene relative abundance based on 16S rRNA gene sequencing we found no difference in any of the genes inolved with acetate, butyrate, or propionate synthesis (P-value `r ifelse(min(pull(picrust_select, bh)) > 0.90, invisible(">"), invisible("<"))` 0.90) [Table S2]. This similarity between groups is highlighted by visualizing genes important in butyrate synthesis [Figure 3A]. We next took a subset of these `r length(metai$sample)` fecal samples and performed metagenomic sequencing to confirm these results. Like the imputed gene results, metagenomic analysis found that there was no difference in any of the genes involved in SCFA synthesis between individuals with normal colons and patients with adenoma or carcinoma (P-value `r ifelse(min(pull(opf_select, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S3]. This lack of difference can be highlighted when we visualize the results for butyrate kinase [Figure 3B]. These observations provide evidence that the gene content does not change due to tumors. Overall, there may be some systematic bias present within our results that may account for the lack of associations.

@@

**Random Forest Models.** The model was first trained on 80% of the data and then tested on the held out 20% (80/20 split) using the Random Forest algorithm for both classification and regression models [@randomforest_citation_2002]. This was repeated on 100 differen 80/20 splits of the data to generate a reasonable range for the AUC of the model. The reported AUCs, unless otherwise specified, are for the test sets. Classification models were built to group normal versus adenoma and normal versus carcinoma or high versus low SCFA concentrations. Regression models were built to assess how well OTUs could be used to get the SCFA concentration within a sample.

**Statistical analysis workflow.** All analysis was performed using the statistical language R [@r_citation_2017]. Generally, differences between the different disease groups used a Kruskal-Walis rank sum test with a Dunn's post-hoc test. Where appropriate Benjamini-Hochberg was used to correct for multiple comparisons [@benjamini_controlling_1995]. First, we assessed differences in SCFA concentrations measured by HPLC between individuals with normal colons and patients with tumors (adenoma or carcinoma). We then analyzed whether SCFA concentrations changed in patients with an adenoma or carcinoma pre- versus post-treatment. Next, we assessed whether OTUs alone or OTUs and SCFAs were better able to classify individuals with and without tumor using Random Forest models. Next, the imputed gene counts of important mediators of SCFA creation were tested. Finally, the counts generated for OPFs that matched important genes involved with SCFA creation were analyzed. Finally, models to classify high or low SCFA concentration based on 16S rRNA gene sequencing data were created using Random Forest. Regression models also were created to classify the exact SCFA concentration based on 16S rRNA gene sequencing data also were built using the Random Forest algorithm. This was done to assess the ability of 16S rRNA gene sequencing to classify the SCFA concentrations.

>>>>

**Decreased SCFA concentrations are not associated with tumors.** We used frozen fecal samples from `r length(metai$sample)` individuals and HPLC to measure acetate, butyrate, and propionate concentrations at a cross-sectional point in time. There was no difference between individuals with normal colons (n=`r length(filter(metai, dx == "normal") %>% pull(sample))`) and patients with either an adenoma (n=`r length(filter(metai, dx == "adenoma") %>% pull(sample))`) or carcinoma (n=`r length(filter(metai, dx == "cancer") %>% pull(sample))`) for any of the SCFAs measured after multiple comparison correction (P-value `r ifelse(min(pull(hplc_groups_scfa_kruskal, bh)) > 0.15, invisible(">"), invisible("<"))` 0.15) [Figure 1A - 1C]. We next measured the concentration of SCFAs in `r length(metaf$initial)` patients with an adenoma (n=`r length(filter(metaf, dx == "adenoma") %>% pull(initial))`) or carcinoma (n=`r length(filter(metaf, dx == "cancer") %>% pull(initial))`) in which we had pre- and post-treatment fecal samples. Although there was a general trend for increasing acetate, butyrate, and propionate concentrations after treatment for tumors, there was no significant difference pre- and post-treatment for patients with adenoma (P-value `r ifelse(min(filter(hplc_treat_scfa_kruskal, disease == "adenoma") %>% pull(bh)) > 0.20, invisible(">"), invisible("<"))` 0.20) or carcinoma (P-value `r ifelse(min(filter(hplc_treat_scfa_kruskal, disease == "carcinoma") %>% pull(bh)) > 0.80, invisible(">"), invisible("<"))` 0.80) [Figure 1D - 1F]. Even though there was no change in SCFA concentrations between individuals with normal colons and those with tumors, this information could still be important to help classify disease.

**Random Forest models with SCFA concentrations do not classify tumors better.** Using the Random Forest algorithm we built models to classify normal versus adenoma and normal versus carcinoma with OTUs or OTUs and SCFA concentrations. For both adenoma and carcinoma models, there was no difference between the median AUC of models with or without SCFA concentrations (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05) [Figure 2]. Although including SCFA concentrations did not add extra information for classification of disease using Random Forest models, it is possible that the genes for enzymes involved in SCFA synthesis may change due to disease status; where a smaller number of microbes are responsible for the observed SCFA concentrations.

**Changes in genes for enzymes involved in SCFA synthesis are not associated with tumors.** Using a list of specific genes that are important for the synthesis of SCFAs [Table S1], we looked for differences in gene abundance based on individuals having normal colons, adenomas, or carcinomas. First, using imputed gene relative abundance based on 16S rRNA gene sequencing we found no difference in any of the genes inolved with acetate, butyrate, or propionate synthesis (P-value `r ifelse(min(pull(picrust_select, bh)) > 0.90, invisible(">"), invisible("<"))` 0.90) [Table S2]. This similarity between groups is highlighted by visualizing genes important in butyrate synthesis [Figure 3A]. Next, we took a subset of these `r length(metai$sample)` fecal samples (n=`r length(geof_samples$X1)`) and performed metagenomic sequencing to confirm these results. Like the imputed gene results, metagenomic analysis found that there was no difference in any of the genes involved in SCFA synthesis between individuals with normal colons (n=`r length(filter(geof_samples, X30 == "Healthy") %>% pull(X2))`) and patients with adenoma (n=`r length(filter(geof_samples, X30 == "Adenoma") %>% pull(X2))`) or carcinoma (n=`r length(filter(geof_samples, X30 == "Cancer") %>% pull(X2))`) (P-value `r ifelse(min(pull(opf_select, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S3]. This lack of difference is highlighted when we visualize the results for butyrate kinase [Figure 3B]. These observations provide evidence that the gene content does not change due to tumors. Although these results suggest that SCFA concentrations do not change due to tumor, there may be some errors present within our results that may account for the lack of associations.

**Expected taxa are associated with higher SCFA concentrations regardless of tumor status.** Using OTU data we built Random Forest models to classify higher than median and lower than median SCFA concentrations. Overall, OTU data had a reasonable ability to classify high and low SCFA concentrations [Figure S1A]. However, these models tended to be overfit, suggesting that rarer taxa may be important for this classification [Figure S1A]. The most important OTUs to these models (assessed with mean decrease in accuracy (MDA)) were to taxa that are normally associated with SCFA production [Figure S1B]. These results highlight that there are no immediately apparent errors in our data because SCFA concentrations are associated with taxa known to produce acetate, butyrate, and propionate. Additionally, OTUs associated with these taxa are the most important to models that can classify high and low SCFA concentrations. Overall, our results do not support the hypothesis that lower SCFA concentrations are associated with colorectal cancer.

@@

**Random Forest Models.** The model was first trained on 80% of the data and then tested on the held out 20% (80/20 split) using the Random Forest algorithm for both classification and regression models [@randomforest_citation_2002]. This was repeated on 100 differen 80/20 splits of the data to generate a reasonable range for the AUC of the model. The reported AUCs, unless otherwise specified, are for the test sets. Classification models were built to group normal versus adenoma and normal versus carcinoma or high versus low SCFA concentrations.

**Statistical analysis workflow.** All analysis was performed using the statistical language R [@r_citation_2017]. Generally, differences between the different disease groups used a Kruskal-Walis rank sum test with a Dunn's post-hoc test. Where appropriate Benjamini-Hochberg was used to correct for multiple comparisons [@benjamini_controlling_1995]. First, we assessed differences in SCFA concentrations measured by HPLC between individuals with normal colons and patients with tumors (adenoma or carcinoma). We then analyzed whether SCFA concentrations changed in patients with an adenoma or carcinoma pre- versus post-treatment. Next, we assessed whether OTUs alone or OTUs and SCFAs were better able to classify individuals with and without tumor using Random Forest models. Next, the imputed gene counts of important mediators of SCFA creation were tested. Finally, the counts generated for OPFs that matched important genes involved with SCFA creation were analyzed. Finally, models to classify high or low SCFA concentration based on 16S rRNA gene sequencing data were created using the Random Forest algorithm.

### 2018-05-22 submission/manuscript.tex
add complete result section template

\textbf{Decreased SCFA concentrations are not associated with tumors.}
We used frozen fecal samples from 490 individuals and HPLC to measure
acetate, butyrate, and propionate concentrations at a cross-sectional
point in time. There was no difference between individuals with normal
colons and patients with either an adenoma or carcinoma for any of the
SCFAs measured after multiple comparison correction (P-value
\textgreater{} 0.15) {[}Figure 1A - 1C{]}. We next measured the
concentration of SCFAs in 67 patients with an adenoma or carcinoma in
which we had pre- and post-treatment fecal samples. Although there was a
general trend for increasing acetate, butyrate, and propionate
concentrations after treatment for tumors, there was no significant
difference pre- and post-treatment for patients with adenoma (P-value
\textgreater{} 0.20) or carcinoma (P-value \textgreater{} 0.80)
{[}Figure 1D - 1F{]}. Even though there was no change in SCFA
concentrations between individuals with normal colons and those with
tumors, this information could still be important to help classify
disease.

\textbf{Random Forest models with SCFA concentrations do not classify
tumors better.} Using the Random Forest algorithm we built models to
classify normal versus adenoma and normal versus carcinoma with OTUs or
OTUs and SCFA concentrations. For both adenoma and carcinoma models,
there was no difference between the median AUC of models with or without
SCFA concentrations (P-value \textgreater{} 0.05) {[}Figure 2{]}.
Although there was no added information gained for classification of
disease by including SCFAs to Random Forest models, it is possible that
the genes for enzymes involved in SCFA pathways may change due to
disease status; where a smaller number of microbes are responsible for
the observed SCFA concentrations.

\textbf{Changes in genes for enzymes involved in SCFA synthesis are not
associated with tumors.} Using a list of specific genes that are
important for the synthesis of SCFAs {[}Table S1{]}, we looked for
differences in gene abundance based on individuals having normal colons,
adenomas, or carcinomas. First, using imputed gene relative abundance
based on 16S rRNA gene sequencing we found no difference in any of the
genes inolved with acetate, butyrate, or propionate synthesis (P-value
\textgreater{} 0.90) {[}Table S2{]}. This similarity between groups is
highlighted by visualizing genes important in butyrate synthesis
{[}Figure 3A{]}. We next took a subset of these 490 fecal samples and
performed metagenomic sequencing to confirm these results. Like the
imputed gene results, metagenomic analysis found that there was no
difference in any of the genes involved in SCFA synthesis between
individuals with normal colons and patients with adenoma or carcinoma
(P-value \textgreater{} 0.70) {[}Table S3{]}. This lack of difference
can be highlighted when we visualize the results for butyrate kinase
{[}Figure 3B{]}. These observations provide evidence that the gene
content does not change due to tumors. Overall, there may be some
systematic bias present within our results that may account for the lack
of associations.

@@

\textbf{Random Forest Models.} The model was first trained on 80\% of
the data and then tested on the held out 20\% (80/20 split) using the
Random Forest algorithm for both classification and regression models
(15). This was repeated on 100 differen 80/20 splits of the data to
generate a reasonable range for the AUC of the model. The reported AUCs,
unless otherwise specified, are for the test sets. Classification models
were built to group normal versus adenoma and normal versus carcinoma or
high versus low SCFA concentrations. Regression models were built to
assess how well OTUs could be used to get the SCFA concentration within
a sample.

\textbf{Statistical analysis workflow.} All analysis was performed using
the statistical language R (16). Generally, differences between the
different disease groups used a Kruskal-Walis rank sum test with a
Dunn's post-hoc test. Where appropriate Benjamini-Hochberg was used to
correct for multiple comparisons (17). First, we assessed differences in
SCFA concentrations measured by HPLC between individuals with normal
colons and patients with tumors (adenoma or carcinoma). We then analyzed
whether SCFA concentrations changed in patients with an adenoma or
carcinoma pre- versus post-treatment. Next, we assessed whether OTUs
alone or OTUs and SCFAs were better able to classify individuals with
and without tumor using Random Forest models. Next, the imputed gene
counts of important mediators of SCFA creation were tested. Finally, the
counts generated for OPFs that matched important genes involved with
SCFA creation were analyzed. Finally, models to classify high or low
SCFA concentration based on 16S rRNA gene sequencing data were created
using Random Forest. Regression models also were created to classify the
exact SCFA concentration based on 16S rRNA gene sequencing data also
were built using the Random Forest algorithm. This was done to assess
the ability of 16S rRNA gene sequencing to classify the SCFA
concentrations.

>>>>

\textbf{Decreased SCFA concentrations are not associated with tumors.}
We used frozen fecal samples from 490 individuals and HPLC to measure
acetate, butyrate, and propionate concentrations at a cross-sectional
point in time. There was no difference between individuals with normal
colons (n=172) and patients with either an adenoma (n=198) or carcinoma
(n=120) for any of the SCFAs measured after multiple comparison
correction (P-value \textgreater{} 0.15) {[}Figure 1A - 1C{]}. We next
measured the concentration of SCFAs in 67 patients with an adenoma
(n=41) or carcinoma (n=26) in which we had pre- and post-treatment fecal
samples. Although there was a general trend for increasing acetate,
butyrate, and propionate concentrations after treatment for tumors,
there was no significant difference pre- and post-treatment for patients
with adenoma (P-value \textgreater{} 0.20) or carcinoma (P-value
\textgreater{} 0.80) {[}Figure 1D - 1F{]}. Even though there was no
change in SCFA concentrations between individuals with normal colons and
those with tumors, this information could still be important to help
classify disease.

\textbf{Random Forest models with SCFA concentrations do not classify
tumors better.} Using the Random Forest algorithm we built models to
classify normal versus adenoma and normal versus carcinoma with OTUs or
OTUs and SCFA concentrations. For both adenoma and carcinoma models,
there was no difference between the median AUC of models with or without
SCFA concentrations (P-value \textgreater{} 0.05) {[}Figure 2{]}.
Although including SCFA concentrations did not add extra information for
classification of disease using Random Forest models, it is possible
that the genes for enzymes involved in SCFA synthesis may change due to
disease status; where a smaller number of microbes are responsible for
the observed SCFA concentrations.

\textbf{Changes in genes for enzymes involved in SCFA synthesis are not
associated with tumors.} Using a list of specific genes that are
important for the synthesis of SCFAs {[}Table S1{]}, we looked for
differences in gene abundance based on individuals having normal colons,
adenomas, or carcinomas. First, using imputed gene relative abundance
based on 16S rRNA gene sequencing we found no difference in any of the
genes inolved with acetate, butyrate, or propionate synthesis (P-value
\textgreater{} 0.90) {[}Table S2{]}. This similarity between groups is
highlighted by visualizing genes important in butyrate synthesis
{[}Figure 3A{]}. Next, we took a subset of these 490 fecal samples
(n=85) and performed metagenomic sequencing to confirm these results.
Like the imputed gene results, metagenomic analysis found that there was
no difference in any of the genes involved in SCFA synthesis between
individuals with normal colons (n=29) and patients with adenoma (n=28)
or carcinoma (n=28) (P-value \textgreater{} 0.70) {[}Table S3{]}. This
lack of difference is highlighted when we visualize the results for
butyrate kinase {[}Figure 3B{]}. These observations provide evidence
that the gene content does not change due to tumors. Although these
results suggest that SCFA concentrations do not change due to tumor,
there may be some errors present within our results that may account for
the lack of associations.

\textbf{Expected taxa are associated with higher SCFA concentrations
regardless of tumor status.} Using OTU data we built Random Forest
models to classify higher than median and lower than median SCFA
concentrations. Overall, OTU data had a reasonable ability to classify
high and low SCFA concentrations {[}Figure S1A{]}. However, these models
tended to be overfit, suggesting that rarer taxa may be important for
this classification {[}Figure S1A{]}. The most important OTUs to these
models (assessed with mean decrease in accuracy (MDA)) were to taxa that
are normally associated with SCFA production {[}Figure S1B{]}. These
results highlight that there are no immediately apparent errors in our
data because SCFA concentrations are associated with taxa known to
produce acetate, butyrate, and propionate. Additionally, OTUs associated
with these taxa are the most important to models that can classify high
and low SCFA concentrations. Overall, our results do not support the
hypothesis that lower SCFA concentrations are associated with colorectal
cancer.

@@

\textbf{Random Forest Models.} The model was first trained on 80\% of
the data and then tested on the held out 20\% (80/20 split) using the
Random Forest algorithm for both classification and regression models
(15). This was repeated on 100 differen 80/20 splits of the data to
generate a reasonable range for the AUC of the model. The reported AUCs,
unless otherwise specified, are for the test sets. Classification models
were built to group normal versus adenoma and normal versus carcinoma or
high versus low SCFA concentrations.

\textbf{Statistical analysis workflow.} All analysis was performed using
the statistical language R (16). Generally, differences between the
different disease groups used a Kruskal-Walis rank sum test with a
Dunn's post-hoc test. Where appropriate Benjamini-Hochberg was used to
correct for multiple comparisons (17). First, we assessed differences in
SCFA concentrations measured by HPLC between individuals with normal
colons and patients with tumors (adenoma or carcinoma). We then analyzed
whether SCFA concentrations changed in patients with an adenoma or
carcinoma pre- versus post-treatment. Next, we assessed whether OTUs
alone or OTUs and SCFAs were better able to classify individuals with
and without tumor using Random Forest models. Next, the imputed gene
counts of important mediators of SCFA creation were tested. Finally, the
counts generated for OPFs that matched important genes involved with
SCFA creation were analyzed. Finally, models to classify high or low
SCFA concentration based on 16S rRNA gene sequencing data were created
using the Random Forest algorithm.
